Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2022-11-04 paper/paper/SPIE/article.tex
[WIP]

@@

% \begin{abstract}
% This document shows the required format and appearance of a manuscript prepared for SPIE journals.
% It is prepared using LaTeX2e with the class file \texttt{spieman.cls}. Please note that the
% following journals require the use of structured abstracts in manuscript submissions:
% \textit{Neurophotonics}, the \textit{Journal of Biomedical Optics}, and the \textit{Journal of
% Medical Imaging}. Structured abstracts are encouraged for the \textit{Journal of
% Micro/Nanolithography, MEMS, and MOEMS}. Guidelines are available on the journal website. Whether
% structured or single-paragraph, the abstract should be a summary of the paper and not an
% introduction. Because the abstract may be used in abstracting and indexing databases, it should be
% self-contained (i.e., no numerical references) and substantive in nature, presenting concisely the
% objectives, methodology used, results obtained, and their significance. A list of up to six
% keywords should immediately follow.
% \end{abstract}

% See https://www.spiedigitallibrary.org/journals/journal-of-medical-imaging/volume-7/issue-01/010101/Integrating-structured-abstracts-in-the-Journal-of-Medical-Imaging/10.1117/1.JMI.7.1.010101.full
% for info on the structured abstract
%
% Purpose: This section presents the significance and aims, stating the broad impact and the
% rationale or motivation of the work, including potentially some background.
%
% Approach: This section briefly describes the materials and methods, including the statistical
% analyses, used in the research.
%
% Results: This section provides a core summary of the findings from the work, including study
% numbers, quantitative analyses, or discoveries.
%
% Conclusions: This section summarizes and interprets the approach and findings from the work,
% stating also the importance or impact of the findings.

% Purpose:
% Data-intensive modeling could provide insight on the broad variability in outcomes in
% spine surgery. Previous studies were limited to analysis of demographic and clinical
% characteristics. We report an analytic framework called “SpineCloud” that incorporates
% quantitative features extracted from perioperative images to predict spine surgery outcome.

% Approach:
% A retrospective study was conducted in which patient demographics, imaging, and outcome
% data were collected. Image features were automatically computed from perioperative CT.
% Postoperative 3- and 12-month functional and pain outcomes were analyzed in terms of improvement
% relative to the preoperative state. A boosted decision tree classifier was trained to predict
% outcome using demographic and image features as predictor variables. Predictions were computed
% based on SpineCloud and conventional demographic models, and features associated with poor outcome
% were identified from weighting terms evident in the boosted tree.

% Results
% Neither approach was predictive of 3- or 12-month outcomes based on preoperative data
% alone in the current, preliminary study. However, SpineCloud predictions incorporating image
% features obtained during and immediately following surgery (i.e., intraoperative and immediate
% postoperative images) exhibited significant improvement in area under the receiver operating
% characteristic (AUC): AUC=0.72 (CI95=0.59 to 0.83) at 3 months and AUC=0.69 (CI95=0.55
% to 0.82) at 12 months.

% Conclusions Predictive modeling of lumbar spine surgery outcomes was improved by incorporation of
% image-based features compared to analysis based on conventional demographic data. The SpineCloud
% framework could improve understanding of factors underlying outcome variability and warrants
% further investigation and validation in a larger patient cohort.
\begin{abstract}

@@

\section*{Results}
% df.groupby(["subgroup", "gross_feature"])["auroc"].median()
Across all classification tasks and analytic choices, we find eiegenfeatures to
more often have predictive utility than not (82.4\% of median AUROCs > 0.5;
median AUROC range across classification tasks 0.47 - 0.64). Simple baseline
reductions on source timeseries, by contrast, were less useful (58.8\% of
median AUROCs > 0.5, median AUROC range across classification tasks 0.42 -
0.62). Additionally, eigenfeature AUROC distributions were overall much more
right-tailed than baseline features suggesting greater predictive potential.
However, performance distributions were wide, and often significantly affected by
analytic choices.

@@

% This document shows the required format and appearance of a manuscript prepared for SPIE journals.
% It is prepared using LaTeX2e with the class file \texttt{spieman.cls}. Please note that the
% following journals require the use of structured abstracts in manuscript submissions:
% \textit{Neurophotonics}, the \textit{Journal of Biomedical Optics}, and the \textit{Journal of
% Medical Imaging}. Structured abstracts are encouraged for the \textit{Journal of
% Micro/Nanolithography, MEMS, and MOEMS}. Guidelines are available on the journal website. Whether
% structured or single-paragraph, the abstract should be a summary of the paper and not an
% introduction. Because the abstract may be used in abstracting and indexing databases, it should be
% self-contained (i.e., no numerical references) and substantive in nature, presenting concisely the
% objectives, methodology used, results obtained, and their significance. A list of up to six keywords
% should immediately follow.
\end{abstract}

@@

\section{Introduction}
\label{sect:intro}  % \label{} allows reference to this section
% This document shows the format and appearance of a manuscript prepared for submission to an SPIE
% journal. Note that this template is only intended to be used as a guideline for author convenience.
% It is designed for optimum clarity and ease of reading for editors and reviewers, but the template
% does not reflect the final page layout of a published journal paper. Accepted papers are
% professionally typeset in XML according to the layout and design of the journal.

@@

\section{Datasets}

@@

In order not to inundate readers with dataset details, and because this is an exploratory
investigation of RMT which is not committed to any specific theory, we refer the reader
to the original publications and/or data releases when greater detail is needed, and
highlight only the most basic aspects of each here.

@@

Dataset scan parameters and acquisition details are summarized in Table \ref{table:1}, and sample
and subgroup sizes are summarized in Table \ref{table:2}. For all datasets, we use the raw
(unpreprocessed) scans, and perform our own preprocessing (detailed later). Given the complexity of
these datasets, we highlight below only those aspects relevant to the current paper.

\subsubsection{Aging}
This dataset\cite{wahlheimIntrinsicFunctionalConnectivity2021} included 34 subjects with a mean age
of 22 years (range: 18-32 years), and 28 subjects with a mean age of 70 years (range: 61-80 years).

% The final sample included 34 younger adults (1832 years old, Mage = 22.21, SD = 3.65; 20 female)
% and 28 older adults (61-80 years old, Mage = 69.82, SD = 5.64; 20 female). Table 1 shows that
% younger and older adults had comparable education and working memory capacity, the latter measured
% by forward and backward digit span (48). Older adults had higher vocabulary scores (49) and slower
% processing speed (50) than younger adults.
% Resting-state data comprised 300 measurements collected over one 10-minute run. This scan duration
% produces reliable functional connectivity estimates within and across participants (57).
% Participants were instructed to remain still and awake with their eyes open. No stimuli appeared
% during this scan, and the display was black. Functional scans were collected using an echo-planar
% image sequence sensitive to blood-oxygen-level-dependent (BOLD) contrast (T2*; 32 slices with 4.0
% mm thickness and no skip, TE = 30 ms, TR = 2000 ms, flip angle = 70, FOV = 220 mm, matrix size =
% 74 × 74 × 32 voxels, A/P phase encoding direction). Slices were collected in a descending order
% that covered the entire cortex and partial cerebellum. At the beginning of the resting-state scan,
% the scanner acquired and discarded two dummy scans.

\subsubsection{Bilinguality}

This dataset\cite{ds001747:1.0.0, goldcarrieelizabethExploringRestingState2018} examined English and
Spanish-speaking monolinguals and multilinguals during a prolonged resting state. The original study
grouped participants into three classes: early vs late bilinguals vs monolingual controls, and found
connectivity differences only between the early and late
bilinguals\cite{goldcarrieelizabethExploringRestingState2018}. We compared for our two classes just
"bilingual" vs "monolingual". The data is most notable for the short TR (875ms) combined with long
acquisition time (12 minutes).

@@

\subsubsection{Learning}

This dataset\cite{schapiroHumanHippocampalReplay2018, schapiroHumanHippocampalReplay2020}. In this study, both rs-fMRI and task fMRI scans
were available for all subjects, allowing for within-subject comparisons. The task is complex and difficult to summarize here adequately, but involved
multiple phases where subjects were presented with various related abstract images, and then later tested for their memory of certain aspects of the learned images. For our study, we simply note that these task scans ought to be broadly more attentionally  and cognitively demanding than the resting state, and instead simply try to classify task vs. test scans using eigenfeatures.

\subsubsection{Dataset: PSYCH}
The PSYCH Dataset  is detailed in \cite{gorgolewskiHighResolution7Tesla2015}. Subjects completed a
battery of psychological measurements, and various resting-state fMRI (rs-fMRI) whole-brain 7-Tesla
scans were obtained. For the purpose of keeping with the ``attention'' theme of the analyses in this
paper, three pairings of subgroups were constructed via z-score splits of the various sum-scores of
scale items involving concepts closely related to attention.

\emph{Trait Attention}.
The PANAS-X \cite{watsonPANASXManualPositive1994} is a 9-point Likert-type self-report measure that
queries the extent to which subjects agree particular affective terms have applied to them over a
recent time period (e.g. a few weeks). An ad hoc (i.e. face-valid) scale was constructed from the
scale items: ``attentive'', ``alert'', ``interested'', ``concentrating'', and reverse-scored items
``sluggish'', ``tired'', ``sleepy'', ``drowsy''. Subjects were then placed in the high / low ``trait
attention'' category based on a median split of the average z-score across items.

\emph{Task Attention}.
The Conjunctive Continuous Performance Task \cite{shalevConjunctiveContinuousPerformance2011} (CCPT)
 is a behavioural measure of sustained attention
where subjects must react only if they see a red square among other different coloured figures. For
this paper, a sum score was constructed by subtracting the number of false positive and false
negative responses from the number of success responses, and subjects with a z-score greater / less
than zero on this scale were placed in high / low ``task attention'' category.

\emph{Vigilance}.
The Mini New York cognition questionnaire assesses various elements of the subject's cognitive
state, e.g. ``my thoughts involved other people'', ``I thought about something positive'' (translated
from German; \cite{gorgolewskiCorrespondenceIndividualDifferences2014}. One question assesses
alertness (``I was fully awake'' / ``war ich vollkommen wach'') and was used to create high / low
``vigilance'' conditions in the same manner as the previous two splits.

\subsubsection{Dataset 2}
Dataset 2 (referred to in figures and tables with the label ``OSTEO'') is detailed in
\cite{tetreaultBrainConnectivityPredicts2016}. Subjects included healthy controls (``nopain''
condition) and individuals with knee osteoarthritis (``pain'' condition) treated for 2 weeks with
either placebo (``pain'' condition) or duloxetine (``duloxetine'' condition). An additional grouping was
also created with both the placebo and duloxetine data (``allpain'' condition), allowing for six
subgroup comparisons. All subjects were imaged with whole-brain rs-fMRI. Scan parameters are
documented in Table \ref{table:1}. The use of rs-fMRI and RMT for examining pain makes this dataset
complementary to \cite{matharooSpontaneousBackpainAlters2020}.

\subsubsection{Dataset 4}
Dataset 4 (referred to in figures and tables with the label ``PARK'' or ``PARKINSONS'') is documented and
described in detail in \cite{madhyasthaDynamicConnectivityRest2015}. Scan parameters are documented
in Table \ref{table:1}. Subjects with non-demented Parkinson's Disease (PD) and healthy controls performed a
number of repetitions of the Attention Network Task
(ANT; \cite{fanActivationAttentionalNetworks2005} ). Subgroup details are available in Table \ref{table:1}. As the ANT
task demands high attention, it is not as straightforward to create ``high'' or ``low'' attention
groups. Instead, since attention presumably is high and varies little in this dataset, and subgroups
instead differ by virtue of the presence or absence of PD, any results here serve as a useful
comparison point for the results from studies that should clearly involve attentional differences.
This dataset also included extensively preprocessed scans, which were included in analyses to
examine the effects of preprocessing on the extracted RMT features.

\subsubsection{Dataset 5}

@@

>>>>

\newcommand{\code}[1]{\small \texttt{#1} \normalsize}

@@

% \begin{abstract} This document shows the required format and appearance of a
% manuscript prepared for SPIE journals. It is prepared using LaTeX2e with the
% class file \texttt{spieman.cls}. Please note that the following journals
% require the use of structured abstracts in manuscript submissions:
% \textit{Neurophotonics}, the \textit{Journal of Biomedical Optics}, and the
% \textit{Journal of Medical Imaging}. Structured abstracts are encouraged for
% the \textit{Journal of Micro/Nanolithography, MEMS, and MOEMS}. Guidelines
% are available on the journal website. Whether structured or single-paragraph,
% the abstract should be a summary of the paper and not an introduction.
% Because the abstract may be used in abstracting and indexing databases, it
% should be self-contained (i.e., no numerical references) and substantive in
% nature, presenting concisely the objectives, methodology used, results
% obtained, and their significance. A list of up to six keywords should
% immediately follow. \end{abstract}

% See
% https://www.spiedigitallibrary.org/journals/journal-of-medical-imaging/volume-7/issue-01/010101/Integrating-structured-abstracts-in-the-Journal-of-Medical-Imaging/10.1117/1.JMI.7.1.010101.full
% for info on the structured abstract
%
% Purpose: This section presents the significance and aims, stating the broad
% impact and the rationale or motivation of the work, including potentially
% some background.
%
% Approach: This section briefly describes the materials and methods, including
% the statistical analyses, used in the research.
%
% Results: This section provides a core summary of the findings from the work,
% including study numbers, quantitative analyses, or discoveries.
%
% Conclusions: This section summarizes and interprets the approach and findings
% from the work, stating also the importance or impact of the findings.

% Purpose: Data-intensive modeling could provide insight on the broad
% variability in outcomes in spine surgery. Previous studies were limited to
% analysis of demographic and clinical characteristics. We report an analytic
% framework called “SpineCloud” that incorporates quantitative features
% extracted from perioperative images to predict spine surgery outcome.

% Approach: A retrospective study was conducted in which patient demographics,
% imaging, and outcome data were collected. Image features were automatically
% computed from perioperative CT. Postoperative 3- and 12-month functional and
% pain outcomes were analyzed in terms of improvement relative to the
% preoperative state. A boosted decision tree classifier was trained to predict
% outcome using demographic and image features as predictor variables.
% Predictions were computed based on SpineCloud and conventional demographic
% models, and features associated with poor outcome were identified from
% weighting terms evident in the boosted tree.

% Results Neither approach was predictive of 3- or 12-month outcomes based on
% preoperative data alone in the current, preliminary study. However,
% SpineCloud predictions incorporating image features obtained during and
% immediately following surgery (i.e., intraoperative and immediate
% postoperative images) exhibited significant improvement in area under the
% receiver operating characteristic (AUC): AUC=0.72 (CI95=0.59 to 0.83) at 3
% months and AUC=0.69 (CI95=0.55 to 0.82) at 12 months.

% Conclusions Predictive modeling of lumbar spine surgery outcomes was improved
% by incorporation of image-based features compared to analysis based on
% conventional demographic data. The SpineCloud framework could improve
% understanding of factors underlying outcome variability and warrants further
% investigation and validation in a larger patient cohort.
\begin{abstract}

@@

\section*{Results}
% df.groupby(["subgroup", "gross_feature"])["auroc"].median()
Across all classification tasks and analytic choices, we find eiegenfeatures to
more often have predictive utility than not (82.4\% of median AUROCs \(>\) 0.5;
median AUROC range across classification tasks 0.47 - 0.64). Simple baseline
reductions on source timeseries, by contrast, were less useful (58.8\% of
median AUROCs \(>\) 0.5, median AUROC range across classification tasks 0.42 -
0.62). Additionally, eigenfeature AUROC distributions were overall much more
right-tailed than baseline features suggesting greater predictive potential.
However, performance distributions were wide, and often significantly affected by
analytic choices.

@@

% This document shows the required format and appearance of a manuscript
% prepared for SPIE journals. It is prepared using LaTeX2e with the class file
% \texttt{spieman.cls}. Please note that the following journals require the use
% of structured abstracts in manuscript submissions: \textit{Neurophotonics},
% the \textit{Journal of Biomedical Optics}, and the \textit{Journal of Medical
% Imaging}. Structured abstracts are encouraged for the \textit{Journal of
% Micro/Nanolithography, MEMS, and MOEMS}. Guidelines are available on the
% journal website. Whether structured or single-paragraph, the abstract should
% be a summary of the paper and not an introduction. Because the abstract may
% be used in abstracting and indexing databases, it should be self-contained
% (i.e., no numerical references) and substantive in nature, presenting
% concisely the objectives, methodology used, results obtained, and their
% significance. A list of up to six keywords should immediately follow.
\end{abstract}

@@

\section{Introduction}
\label{sect:intro}  % \label{} allows reference to this section
% This document shows the format and appearance of a manuscript prepared for
% submission to an SPIE journal. Note that this template is only intended to be
% used as a guideline for author convenience. It is designed for optimum
% clarity and ease of reading for editors and reviewers, but the template does
% not reflect the final page layout of a published journal paper. Accepted
% papers are professionally typeset in XML according to the layout and design
% of the journal.

@@

\section{Datasets}
\label{sec:datasets}

@@

@@

In order not to inundate readers with dataset details, and because this is an
exploratory investigation of RMT which is not committed to any specific
theory\footnote{In fact, datasets and, later, the classification tasks, were
chosen without investigation into any findings or publications associated with
the data, both in order to reduce bias in our decisions, and because the
original authors analytic / statistical approaches and intentions have very
limited relevance to our whole-brain, voxelwise, multiverse predictive
approach.}, we refer the reader to the original publications and/or data
releases when greater detail is needed, and highlight only the most basic
aspects of each dataset here.

Dataset scan parameters and acquisition details are summarized in Table
\ref{table:1}, and sample and subgroup sizes are summarized in Table
\ref{table:2}.

\subsection{Datasets and Classification Tasks}

The \textit{Aging} dataset\cite{wahlheimIntrinsicFunctionalConnectivity2021}
included rs-fMRI data for 34 subjects with a mean age of 22 years (range: 18-32
years), and 28 subjects with a mean age of 70 years (range: 61-80 years). For
our analysis, we make the binary classification task the prediction of subject
age-group membership, i.e. \code{younger v older}.

% The final sample included 34 younger adults (1832 years old, Mage = 22.21, SD
% = 3.65; 20 female) and 28 older adults (61-80 years old, Mage = 69.82, SD =
% 5.64; 20 female). Table 1 shows that younger and older adults had comparable
% education and working memory capacity, the latter measured by forward and
% backward digit span (48). Older adults had higher vocabulary scores (49) and
% slower processing speed (50) than younger adults. Resting-state data
% comprised 300 measurements collected over one 10-minute run. This scan
% duration produces reliable functional connectivity estimates within and
% across participants (57). Participants were instructed to remain still and
% awake with their eyes open. No stimuli appeared during this scan, and the
% display was black. Functional scans were collected using an echo-planar image
% sequence sensitive to blood-oxygen-level-dependent (BOLD) contrast (T2*; 32
% slices with 4.0 mm thickness and no skip, TE = 30 ms, TR = 2000 ms, flip
% angle = 70, FOV = 220 mm, matrix size = 74 × 74 × 32 voxels, A/P phase
% encoding direction). Slices were collected in a descending order that covered
% the entire cortex and partial cerebellum. At the beginning of the
% resting-state scan, the scanner acquired and discarded two dummy scans.

\subsubsection{Bilinguality Dataset}

The \textit{Bilinguality} dataset\cite{ds001747:1.0.0,
goldcarrieelizabethExploringRestingState2018} examined English and
Spanish-speaking monolinguals and multilinguals during a prolonged resting
state. The original study grouped participants into three subgroups: early vs
late bilinguals vs monolingual
controls\cite{goldcarrieelizabethExploringRestingState2018}. For our analysis,
we make the binary classification task be to classify \code{monolingual v
bilingual}.

@@

\subsubsection{Depression Dataset}

\subsubsection{Learning Dataset}

The \textit{Learning} dataset\cite{schapiroHumanHippocampalReplay2018,
schapiroHumanHippocampalReplay2020} has both rs-fMRI and task fMRI scans were
available for all subjects, and \code{task v rest} is chosen as the binary
classification task for our analysis.  The task is complex and difficult to
summarize here adequately, but involved multiple phases where subjects were
presented with various related abstract images, and then later tested for their
memory of certain aspects of the learned images. Thus the classification task
might also be considered learning vs. rest.

\subsubsection{Attention Dataset}

The \textit{Attention} dataset\cite{gorgolewskiHighResolution7Tesla2015} is a
high-resolution rs-fMRI dataset including a battery of psychological measures
for each subject. Since previous studies employing RMT sometimes interpreted
their findings with respect to attentional processes
\cite{wangRandomMatrixTheory2016,matharooSpontaneousBackpainAlters2020}, we
divided subjects into various high vs. low attention binary classification
tasks based on median splits of subsets of the metrics available in this study.

The \code{viglant v nonvigilant} task was formed based on self-report
questionnaire items involving
"vigilance"\cite{gorgolewskiCorrespondenceIndividualDifferences2014}. The
\code{trait\_attend v trait\_nonattend} task was formed using PANAS-X
\cite{watsonPANASXManualPositive1994} items related to self-reported
wakefulness and attention over the past weeks. Finally, the
\code{task\_attend v tas\_nonattent} classification task was constructed
using the scores of The Conjunctive Continuous Performance Task
(CCPT\cite{shalevConjunctiveContinuousPerformance2011}). This is a behavioural
performance task which requires the subject to quickly and selectively respond
to only certain visual stimuli. Details allowing exact reproduction of these
splits are available in this paper's
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/preprocess/unify\_attention\_data.py}{source
code}.

\subsubsection{Osteoarthritis Dataset}
The \textit{Osteo} dataset\cite{tetreaultBrainConnectivityPredicts2016}
includes whiole-brain rs-fMRI scans of healthy, pain-free controls
(\code{nopain} condition), and individuals with knee osteoarthritis.
Osteoarthritic patients were treated for 2 weeks with either placebo
(\code{pain} condition) or duloxetine (\code{duloxetine} condition).
We use the three binary classification tasks \code{nopain v duloxetine},
\code{nopain v pain}, and \code{pain v duloxetine} for our analysis.

\subsubsection{Parkinson's Dataset}
The \textit{Park} dataset\cite{madhyasthaDynamicConnectivityRest2015} includes
subjects with non-demented Parkinson's Disease (PD) and
healthy controls performed a number of repetitions of the Attention Network
Task (ANT; \cite{fanActivationAttentionalNetworks2005}) during scans.
The classification task for this dataset is Parkinson's versus controls, i.e.
\code{park v ctrl}.

@@

\subsection{Classification Task Subgroups}

For each dataset, we construct subgroups based either on existing labels in the
data, or for the \textit{Attention} dataset, from psychological measures available
from the data.
